---
title: "Visualizing `RangeSummarizedExperiment` objects Shiny Apps using epivizrChart"
author: "Jayaram Kancherla, Hector Corrada Bravo"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Visualizing `RangeSummarizedExperiment` objects Shiny Apps using epivizrChart}
  %\usepackage[UTF-8]{inputenc}
---

In this vignette, we will build a shiny app to visualize `RangeSummarizedExperiment` using epivizrChart. Since epiviz visualization library is built upon the web components framework, it can be integrated with most frameworks that support HTML. 

```{r setup, eval=TRUE, include=FALSE}
library(epivizrChart)
library(shiny)
library(Homo.sapiens)
```

Sample data sets to use for the vignette.

```{r}
data(sumexp)
```

We create an Environment element which visualizes genome wide data. We then visualize `cancer` and `normal` values from the `SummarizedExperiment` object.

```{r}
epivizEnv <- epivizEnv(interactive = TRUE)
scatterplot <- epivizEnv$plot(sumexp, datasource_name="sumExp", columns=c("cancer", "normal"))
```

After looking at the genomic wide data, if you are interested in further exploring a specific region of the genome, We can create a navigation element linked to that genomic location. We can plot additional annotation/data charts/tracks in this region. 

```{r}
epivizNav <- epivizNav(chr="chr11", start=118000000, end=121000000, parent=epivizEnv, interactive = TRUE)

genes_track <- epivizNav$add_genome(Homo.sapiens, datasource_name="genes")
region_scatterplot <- epivizNav$plot(sumexp, datasource_name="sumExp", columns=c("normal", "cancer"))
```

Finally, we can embed these components in a Shiny App.

```{r}

app <- shinyApp(
  ui=fluidPage(
    uiOutput("epivizChart")
  ),
  server=function(input, output, session) {
    
    output$epivizChart <- renderUI({
      epivizEnv$render_component(shiny=TRUE)
    })
    
    epivizEnv$register_shiny_handler(session)
  }
)

app

```